---
title: "week 8 hw"
author: "Roxana Nojoumi"
date: "March 5, 2017"
output: md_document
---
#PART 1
```{r}
#install.packages("paleotree")
library(ggplot2)
canids <- read.csv("~/Desktop/eeb-177/homework/canids_min_max.csv", header = F, as.is = T)
names(canids) <- c("genus", "species", "minage", "maxage")
head(canids)

canid_occ <- ggplot(canids, aes( species, ymin = maxage, ymax=minage, colour = genus))
canid_occ <- canid_occ + geom_linerange()+ theme(legend.position="none")+ coord_flip()+  theme(axis.text.y = element_text(size=3))+ theme(axis.ticks.y=element_blank()) + scale_y_continuous(limits=c(0, 40), expand = c(0, 0), breaks=c(0, 10, 20, 30, 40)) + labs(title = "Canid Fossil Occurrences", x = "Species", y = "Ma ago") + theme(plot.title = element_text(hjust = 0.5, size=22, face = "bold"), axis.title =element_text(size=20))




canid_occ

ggsave(filename = "canid-occ.pdf", plot = canid_occ)




```

1.What taxonomic group is displayed?
The figure is for the group Canids. 

2.How many unique fossil species are contained within it?
cut -d "," -f 2 canids_min_max.csv | sort|uniq| wc -l
160



3.How many unique fossil genera?
cut -d "," -f 1 canids_min_max.csv |sort| uniq| wc -l
58

4.What species has the longest(range)?
Specie with the longest fossil range is:  Cynarctoides acridens and its fossil range is:  [17.200000000000003]






#For family equidae
```{r}
library(ggplot2)
equidaes <- read.csv("~/Desktop/eeb-177/homework/equidae_min_max.csv", header = F, as.is = T)
names(equidaes) <- c("genus", "species", "minage", "maxage")
head(equidaes)

equidaes_occ <- ggplot(equidaes, aes( species, ymin = maxage, ymax=minage, colour = genus))
equidaes_occ <- equidaes_occ + geom_linerange()+ theme(legend.position="none")+ coord_flip()+  theme(axis.text.y = element_text(size=3))+ theme(axis.ticks.y=element_blank()) + scale_y_continuous(limits=c(0, 40), expand = c(0, 0), breaks=c(0, 10, 20, 30, 40)) + labs(title = "Equidae Fossil Occurrences", x = "Species", y = "Ma ago") + theme(plot.title = element_text(hjust = 0.5, size=22, face = "bold"), axis.title =element_text(size=20))




equidaes_occ

ggsave(filename = "equidaes-occ.pdf", plot = equidaes_occ)


```


1.What taxonomic group is displayed?
The figure is for the group Equidae. 

2.How many unique fossil species are contained within it?
cut -d "," -f 2 equidae_min_max.csv | sort|uniq| wc -l
295



3.How many unique fossil genera?
cut -d "," -f 1 equidae_min_max.csv |sort| uniq| wc -l
53

4.What species has the longest(range)?
Specie with the longest fossil range is:  Hipparion aff. crassum and its fossil range is:  [23.0183]


#EXTRA CREDT
stratigraphic ranges of each genus
```{r}
library(ggplot2)
canids <- read.csv("~/Desktop/eeb-177/homework/canids-genus-ranges.csv", header = F, as.is = T)
names(canids) <- c("genus", "minage", "maxage")
head(canids)

canid_occ <- ggplot(canids, aes( genus, ymin = maxage, ymax=minage, colour = genus))
canid_occ <- canid_occ + geom_linerange()+ theme(legend.position="none")+ coord_flip()+  theme(axis.text.y = element_text(size=3))+ theme(axis.ticks.y=element_blank()) + scale_y_continuous(limits=c(0, 40), expand = c(0, 0), breaks=c(0, 10, 20, 30, 40)) + labs(title = "Canid Fossil Occurrences", x = "Genera", y = "Ma ago") + theme(plot.title = element_text(hjust = 0.5, size=22, face = "bold"), axis.title =element_text(size=20))

canid_occ


```






#PART 2: Data Frames

#1- Exploring data frames


#EXAMPLES
```{r}
cats <- read.csv(file = "~/Desktop/eeb-177/homework/feline-data.csv")
#Adding a column
age <- c(4,5,8)
cats <- cbind(cats, age)

#adding a row
#newRow <- list("tortoiseshell", 3.3, TRUE, 9)
#cats <- rbind(cats, newRow)
#get error because tortoiseshell has to be a level
levels(cats$coat) <- c(levels(cats$coat), 'tortoiseshell')
cats <- rbind(cats, list("tortoiseshell", 3.3, TRUE, 9))
#cats
#change a factor column to a character vector
cats$coat <- as.character(cats$coat)
#str(cats)
#cats
#remove a row, this removes the netire 4th row
cats[-4,]

#if want to omit all rows with Na
cats <- na.omit(cats)

#adding two rows(lists)
#cats <- rbind(cats, cats)
#cats

#to remove row names and so R rename them sequentially 
rownames(cats) <- NULL
cats <- rbind(cats, cats)
cats

```
# CHALLENGE 1
Make a data frame that holds the following information for yourself:

first name
last name
lucky numbe
Then use rbind to add an entry for the people sitting beside you. Finally, use cbind to add a column with each person’s answer to the question, “Is it time for coffee break?”

```{r}
df <- data.frame(first = c('Roxana'),
                 last = c('Nojoumi'),
                 lucky_number = c(7),
                 stringsAsFactors = FALSE)
df <- rbind(df, list('Mehdi', 'Nojoumi', 7) )
df <- cbind(df, coffeetime = c(FALSE,TRUE))
df
```
# Examples

To download a file from internet then read it
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv", destfile = "data/gapminder-FiveYearData.csv")
gapminder <- read.csv("data/gapminder-FiveYearData.csv")
to download and read file directly
```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv")
str(gapminder)
```
#examine individual column
```{r}
typeof(gapminder$year)
typeof(gapminder$country)
str(gapminder$country)
#interrogate about dimension
length(gapminder)
#to get number of rows or columns
nrow(gapminder)
ncol(gapminder)
#or both at once
dim(gapminder)
#to see title of the columns
colnames(gapminder)
#to see first few lines
head(gapminder)
```
# CHALLENGE 2 
Go to file -> new file -> R script, and write an R script to load in the gapminder dataset. Put it in the scripts/ directory and add it to version control.
Run the script using the source function, using the file path as its argument (or by pressing the “source” button in RStudio).
```{r}
#content of the script
#download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pa#ges/_episodes_rmd/data/gapminder-FiveYearData.csv", destfile = #"data/gapminder-FiveYearData.csv")
#gapminder <- read.csv(file = "data/gapminder-FiveYearData.csv")



#source(file = "~/scripts/load-gapminder.R")
```

#CHALLENGE 3
Read the output of str(gapminder) again; this time, use what you’ve learned about factors, lists and vectors, as well as the output of functions like colnames and dim to explain what everything that str prints out for gapminder means. If there are any parts you can’t interpret, discuss with your neighbors!


answer
The object gapminder is a data frame with columns
country and continent are factors.
year is an integer vector.
pop, lifeExp, and gdpPercap are numeric vectors.


#2 Subsetting Data 

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
x


#Accessing elements with indices

x[1]
x[4]
#2 elements or more
x[c(1, 3)]
x[1:4]
1:4
c(1, 2, 3, 4)
x[c(1,1,3)]
x[6]
x[0]
```


# Skipping and removing elements
```{r}
#exclude and element
x[-2]
x[c(-1, -5)]  # or x[-c(1,5)]
#minus sign before parantecis
x[-(1:3)]
#to remove a vector, assign it 
x <- x[-4]
x
```

# Challenge 1
Given the following code
```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```
Come up with at least 3 different commands that will produce the following output:

  b   c   d 
6.2 7.1 4.8 


#Answer
```{r}
#1
x[2:4]

#2
x[-c(1,5)]

#3
x[c("b", "c", "d")]

#4
x[c(2,3,4)]

```


#Subsetting by name
```{r}
#extract elements by name
x[c("a", "c")]
#to skip or remove
x[-which(names(x) == "a")]
names(x) == "a"
which(names(x) == "a")
#skipping multiple names
x[-which(names(x) %in% c("a", "c"))]
```


# Challenge 2:
Run the following code to define vector x as above
```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```
Given this vector x, what would you expect the following to do?
```{r}
x[-which(names(x) == "g")]
```
Try out this command and see what you get. Did this match your expectation? Why did we get this result? (Tip: test out each part of the command on it’s own - this is a useful debugging strategy)

Which of the following are true:

A) if there are no TRUE values passed to which, an empty vector is returned
B) if there are no TRUE values passed to which, an error message is shown
C) integer() is an empty vector
D) making an empty vector negative produces an “everything” vector
E) x[] gives the same result as x[integer()]

#Answer
A and C are correct.
The which command returns the index of every TRUE value in its input. The names(x) == "g" command didn’t return any TRUE values. Because there were no TRUE values passed to the which command, it returned an empty vector. Negating this vector with the minus sign didn’t change its meaning. Because we used this empty vector to retrieve values from x, it produced an empty numeric vector. It was a named numeric empty vector because the vector type of x is “named numeric” since we assigned names to the values (try str(x) ).

```{r}
x <- 1:3
x

names(x) <- c('a', 'a', 'a')
x

x['a']  # only returns first value

x[which(names(x) == 'a')]  # returns all three values

names(x) == c('a', 'c')

names(x) == c('a', 'c', 'e')
```


#Subsetting through other logical operations
```{r}
x[c(TRUE, TRUE, FALSE, FALSE)]

x[c(TRUE, FALSE)]

x[x > 7]
```

#Challenge 3

Given the following code
```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```
Write a subsetting command to return the values in x that are greater than 4 and less than 7.
#Answer
```{r}
subset <- x[x<7 & x>4]
print(subset)
```

#Handling special values
#Factor subsetting
```{r}
f <- factor(c("a", "a", "b", "c", "c", "d"))
f[f == "a"]

f[f %in% c("b", "c")]

f[1:3]

f[-3]
```


#Matrix subsetting
```{r}
set.seed(1)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m[3:4, c(3,1)]

m[, c(3,4)]

m[3,]

m[3, , drop=FALSE]

#m[, c(3,6)] #supposed to give an error!

m[5]

matrix(1:6, nrow=2, ncol=3)

matrix(1:6, nrow=2, ncol=3, byrow=TRUE)
```



#Challenge 4

Given the following code
```{r}
m <- matrix(1:18, nrow=3, ncol=6)
print(m)
```

Which of the following commands will extract the values 11 and 14?

A. m[2,4,2,5]

B. m[2:5]

C. m[4:5,2]

D. m[2,c(4,5)]

Answer
D

```{r}
#D. 
m[2,c(4,5)]
```


#List subsetting
```{r}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
xlist[1]

xlist[1:2]

xlist[[1]]

#xlist[[1:2]] #supposed to get an error!
#xlist[[-1]]

xlist[["a"]]
xlist$data
```

#Challenge 5
Given the following list
```{r}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))

```

Using your knowledge of both list and vector subsetting, extract the number 2 from xlist. Hint: the number 2 is contained within the “b” item in the list.

#Answer
```{r}
xlist$b[2]
#OR
xlist[[2]][2]
#OR
xlist[["b"]][2]
```

# Challenge 6
Given a linear model
```{r}
mod <- aov(pop ~ lifeExp, data=gapminder)
```

Extract the residual degrees of freedom (hint: attributes() will help you)

#Answer
```{r}
attributes(mod) ## `df.residual` is one of the names of `mod`
mod$df.residual
```

#Data frames
```{r}
head(gapminder[3])

head(gapminder[["lifeExp"]])

head(gapminder$year)

gapminder[1:3,]

gapminder[3,]
```


# Challenge 7
Fix each of the following common data frame subsetting errors:

<br><br>

#Answer
1-Extract observations collected for the year 1957
```{r}
# gapminder[gapminder$year = 1957,]
gapminder[gapminder$year == 1957,]
```


2-Extract all columns except 1 through to 4
```{r}
# gapminder[,-1:4]
gapminder[,-c(1:4)]
```

3-Extract the rows where the life expectancy is longer the 80 years
```{r}
# gapminder[gapminder$lifeExp > 80]
gapminder[gapminder$lifeExp > 80,]
```

4-Extract the first row, and the fourth and fifth columns (lifeExp and gdpPercap).
```{r}
# gapminder[1, 4, 5]
gapminder[1, c(4, 5)]
```

5-Advanced: extract rows that contain information for the years 2002 and 2007
```{r}
# gapminder[gapminder$year == 2002 | 2007,]
 gapminder[gapminder$year == 2002 | gapminder$year == 2007,]
 gapminder[gapminder$year %in% c(2002, 2007),]
```

# Challenge 8

1-Why does gapminder[1:20] return an error? How does it differ from gapminder[1:20, ]?
#Answer
gapminder is a data frame so needs to be subsetted on two dimensions. gapminder[1:20, ] subsets the data to give the first 20 rows and all columns.
<br><br>
2-Create a new data.frame called gapminder_small that only contains rows 1 through 9 and 19 through 23. You can do this in one or two steps.
#Asnwer
```{r}
gapminder_small <- gapminder[c(1:9, 19:23),]
gapminder_small
```
#Creating Publication-Quality Graphics

```{r}
library("ggplot2")
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

#Challenge 1
Modify the example so that the figure visualise how life expectancy has changed over time:
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) + geom_point()
Hint: the gapminder dataset has a column called “year”, which should appear on the x-axis.
#Answer
```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp)) + geom_point()
```
#Challenge 2
In the previous examples and challenge we’ve used the aes function to tell the scatterplot geom about the x and y locations of each point. Another aesthetic property we can modify is the point color. Modify the code from the previous challenge to color the points by the “continent” column. What trends do you see in the data? Are they what you expected?
<br><br>
#Answer
```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp, color=continent)) +
  geom_point()
```
<br><br>
Layers

```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line()
```
```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line() + geom_point()

```
```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
  geom_line(aes(color=continent)) + geom_point()
```
#Challenge 3
Switch the order of the point and line layers from the previous example. What happened?
<br><br>

#Answer
The lines now get drawn over the points!

```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
 geom_point() + geom_line(aes(color=continent))
```

#Transformations and statistics
```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color=continent)) +
  geom_point()
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5) + scale_x_log10()
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm")
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm", size=1.5)
```

#Challenge 4a
Modify the color and size of the points on the point layer in the previous example.

Hint: do not use the aes function.
<br><br>
#Answer
```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
 geom_point(size=3.5, color="purple") + scale_x_log10() +
 geom_smooth(method="lm", size=1.5)
```

#Challenge 4b
Modify your solution to Challenge 4a so that the points are now a different shape and are colored by continent with new trendlines. Hint: The color argument can be used inside the aesthetic.
```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
geom_point(size=3, pch=17) + scale_x_log10() +
geom_smooth(method="lm", size=1.5)
```
<br><br>
#Multi-panel figures
```{r}
starts.with <- substr(gapminder$country, start = 1, stop = 1)
az.countries <- gapminder[starts.with %in% c("A", "Z"), ]
ggplot(data = az.countries, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country)
```
<br><br>
#Modifying text
```{r}
ggplot(data = az.countries, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  xlab("Year") + ylab("Life expectancy") + ggtitle("Figure 1") +
  scale_colour_discrete(name="Continent") +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())
```

#Challenge 5
Create a density plot of GDP per capita, filled by continent.

Advanced:

Transform the x axis to better visualise the data spread.
Add a facet layer to panel the density plots by year.
<br><br>
#Answer
```{r}
ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
 geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()
```



